Contracts for Contracts

### 2023-06-07 paper/contract_solidity.tex
paper
misc
update sec 3

\section{Background: The Solidity Language}

brief introduction of the Solidity language, why it is complicated

@@

In this section, we introduce \lang (Contract Solidity) informally with examples.
\lang is a non-breaking extension that additionally provides means
to specify behavioral contracts on top of the Solidity language.
We provide a brief introduction of Solidity first, then discuss our extension
of behavioral contracts.
We often use \emph{specification} for behavioral contracts to avoid terminological ambiguity.

\paragraph{\lang's Design Rationale}
This paper concerns about designing an extension for Solidity that enhances the
language with higher-order temporal behavioral contracts.
The extension should be
\todo{non-intrusive, easy to use, expressive, gradual (you do not need to pay the overhead
	if you don't use contracts), efficient (little overhead)}

@@

\subsubsection*{Limitations}
Address contracts are \emph{not} higher-order -- if an address takes another
address as argument, then we cannot enforce the contract of the argument
address.
For example, in the following case, @b@ itself has an ``address contract''
@<b-pre> -> <b-post>@ but there is no general way to enforce that.
Because we do not have the control of the actual callee function of @a@,
therefore no way to modify the code of @a@ to check the pre-condition
and post-condition when @b@ is called in @a@.

@@

For example, the following identity function of addresses has a spec
for the argument address. But the body of function does not invoke
the address, instead, it directly returns @a@ to the caller.
Once address @a@ escapes, we cannot enforce the condition anymore.

@@

\todo{where to add message sender? proposed syntax:}
\begin{lstlisting}
  # ts1 @ f(x) returns z from sdr1 => ts2 @ g(a) returns c from sdr2
\end{lstlisting}
\begin{lstlisting}
  # f(x) returns z => g(a) returns c /\ f.msg.sender == g.msg.sender
\end{lstlisting}
\wac{\texttt{msg.sender} is better tied with transaction. We can use ts1.msg.sender and I prefer to put the constraints in the side-cond. }

@@

Positive temporal properties enforces that an event must happen
after another event. Syntax:
\begin{lstlisting}
  # ts1 @ f(x) returns z => ts2 @ g(a) returns c /\ side-cond
\end{lstlisting}
It enforces that when @g@ is invoked, we check @f@ has invoked \emph{and}
the side-condition is evaluated to true.
We do not enforce that @g@ will be invoked eventually after invoking @f@.

Side conditions can be arbitrary expressions that uses the time-stamps, message
sender, arguments, return values of the two function calls.
For example, the following spec passes arguments of @f@ and @g@ into
another function for checking:

\begin{lstlisting}
  # ts1 @ f(x) returns z => ts2 @ g(a) returns c /\ check(x, a)
\end{lstlisting}

If a variable occurs in both events, an additional equality constraint
is synthesized.  For example,
\begin{lstlisting}
  # ts1 @ f(x) returns z => ts1 @ g(z) returns c
\end{lstlisting}
requires that both function @f@ and @g@ happen in the same transaction,
and the return value of @f@ is the same as the argument of @g@.
It will be translated to the core syntax as the following:
\begin{lstlisting}
  # ts1 @ f(x) returns z => ts2 @ g(y) returns c /\ ts1 == ts2 && x == y
\end{lstlisting}

It is totally fine to omit the time-stamp, arguments, or return values.
Simply requiring @f@ happens before @g@:
\begin{lstlisting}
  # f => g
  function f(...) { ... }
\end{lstlisting}

@@

Note: @f => g@ matches the most recent call of @f@,
or any prior call of @f@ (if there are multiple calls of @f@)?

@@

\begin{lstlisting}
  # ts1 @ f(x) returns z =/> ts2 @ g(a) returns c /\ side-cond
\end{lstlisting}
It enforces that when @g@ is invoked \emph{and} the side condition
evaluates to true, we check that @f@ has not been invoked.

We do not enforce that @g@ cannot be invoked eventually
after invoking @f@.

For example, the following spec enforces that @g@ cannot be
invoked after @f@ using @z@ as the argument:
\begin{lstlisting}
  # ts1 @ f(x) returns z =/> ts2 @ g(z) returns c
  //or equivalently
  # ts1 @ f(x) returns z =/> ts2 @ g(y) returns c /\ x == y
\end{lstlisting}
However it is okay to call @g@ after @f@ with an argument different from @z@.

@@

Using negative temporal contracts, we can enforce affine use, i.e.
a function can be called at most once.
\begin{lstlisting}
  # f =/> f
  function f(int x) { ... }
\end{lstlisting}
If @f@ has been called before (and the call is finished), any
call to @f@ afterwards is prohibited.

@@

For example, the following spec enforces that @g@ can only be invoked
under the calling context of @h@, and the return value of @h@ must be
the same as the argument of @g@:

@@

To enforce this behavioral contracts, we must check
\begin{itemize}
	\item when @g@ is invoked, there is a frame of @h@ on the stack,
	\item when @h@ returns (which happens after @g@ returns), the return value
	      is the same as the argument of @g@.
\end{itemize}

Question: what if @g@ is invoked multiple times under the same calling context of
@h@?

Solution: need quantifier. A simple/default solution is to check all
invocations of @g@.

Question: what if there are multiple call frames of @h@ on the stack?

Solution: need quantifier. A simple/default solution is to only check
the most recent call frame of @h@ on the stack.

>>>>

@@

We now introduce \lang informally with examples.
\lang is a non-breaking extension that additionally provides means
to specify behavioral contracts on top of the Solidity language.
We provide a brief introduction of Solidity first, then discuss our extension
of behavioral contracts.
We often use \emph{specification} for behavioral contracts to avoid terminological ambiguity.

\paragraph{Design Rationale}
This paper concerns about designing an extension for Solidity that enhances the
language with higher-order temporal behavioral contracts.
The extension should be
\todo{non-intrusive, easy to use, expressive, gradual (you do not need to pay the overhead
	if you don't use contracts), efficient (little overhead)}

@@

\subsubsection*{Limitations}
Address contracts are \emph{not} higher-order -- if an address takes another
address as argument, then we cannot enforce the contract of the argument
address.
For example, in the following case, \code{b} itself has an ``address contract''
\code{<b-pre> -> <b-post>} but there is no general way to enforce that.
Because we do not have the control of the actual callee function of \code{a},
therefore no way to modify the code of \code{a} to check the pre-condition
and post-condition when \code{b} is called in \code{a}.

@@

For example, the following identity function of addresses has a spec
for the argument address. But the body of function does not invoke
the address, instead, it directly returns \code{a} to the caller.
Once address \code{a} escapes, we cannot enforce the condition anymore.

@@

\todo{where to add message sender? proposed syntax:}
\begin{lstlisting}
  # ts1 \code{ f(x) returns z from sdr1 => ts2 } g(a) returns c from sdr2
\end{lstlisting}
\begin{lstlisting}
  # f(x) returns z => g(a) returns c /\ f.msg.sender == g.msg.sender
\end{lstlisting}
\wac{\texttt{msg.sender} is better tied with transaction. We can use ts1.msg.sender and I prefer to put the constraints in the side-cond. }

@@

Positive temporal properties enforces that an event must happen
after another event. Syntax:
\begin{lstlisting}
  # ts1 \code{ f(x) returns z => ts2 } g(a) returns c /\ side-cond
\end{lstlisting}
It enforces that when \code{g} is invoked, we check \code{f} has invoked \emph{and}
the side-condition is evaluated to true.
We do not enforce that \code{g} will be invoked eventually after invoking \code{f}.

Side conditions can be arbitrary expressions that uses the time-stamps, message
sender, arguments, return values of the two function calls.
For example, the following spec passes arguments of \code{f} and \code{g} into
another function for checking:

\begin{lstlisting}
  # ts1 \code{ f(x) returns z => ts2 } g(a) returns c /\ check(x, a)
\end{lstlisting}

If a variable occurs in both events, an additional equality constraint
is synthesized.  For example,
\begin{lstlisting}
  # ts1 \code{ f(x) returns z => ts1 } g(z) returns c
\end{lstlisting}
requires that both function \code{f} and \code{g} happen in the same transaction,
and the return value of \code{f} is the same as the argument of \code{g}.
It will be translated to the core syntax as the following:
\begin{lstlisting}
  # ts1 \code{ f(x) returns z => ts2 } g(y) returns c /\ ts1 == ts2 && x == y
\end{lstlisting}

It is totally fine to omit the time-stamp, arguments, or return values.
Simply requiring \code{f} happens before \code{g}:
\begin{lstlisting}
  # f => g
  function f(...) { ... }
\end{lstlisting}

@@

Note: \code{f => g} matches the most recent call of \code{f},
or any prior call of \code{f} (if there are multiple calls of \code{f})?

@@

\begin{lstlisting}
  # ts1 \code{ f(x) returns z =/> ts2 } g(a) returns c /\ side-cond
\end{lstlisting}
It enforces that when \code{g} is invoked \emph{and} the side condition
evaluates to true, we check that \code{f} has not been invoked.

We do not enforce that \code{g} cannot be invoked eventually
after invoking \code{f}.

For example, the following spec enforces that \code{g} cannot be
invoked after \code{f} using \code{z} as the argument:
\begin{lstlisting}
  # ts1 \code{ f(x) returns z =/> ts2 } g(z) returns c
  //or equivalently
  # ts1 \code{ f(x) returns z =/> ts2 } g(y) returns c /\ x == y
\end{lstlisting}
However it is okay to call \code{g} after \code{f} with an argument different from \code{z}.

@@

Using negative temporal contracts, we can enforce affine use, i.e.
a function can be called at most once.
\begin{lstlisting}
  # f =/> f
  function f(int x) { ... }
\end{lstlisting}
If \code{f} has been called before (and the call is finished), any
call to \code{f} afterwards is prohibited.

@@

For example, the following spec enforces that \code{g} can only be invoked
under the calling context of \code{h}, and the return value of \code{h} must be
the same as the argument of \code{g}:

@@

To enforce this behavioral contracts, we must check
\begin{itemize}
	\item when \code{g} is invoked, there is a frame of \code{h} on the stack,
	\item when \code{h} returns (which happens after \code{g} returns), the return value
	      is the same as the argument of \code{g}.
\end{itemize}

Question: what if \code{g} is invoked multiple times under the same calling context of
\code{h}?

Solution: need quantifier. A simple/default solution is to check all
invocations of \code{g}.

Question: what if there are multiple call frames of \code{h} on the stack?

Solution: need quantifier. A simple/default solution is to only check
the most recent call frame of \code{h} on the stack.
